A Partitioning Algorithm for Maximum Common Subgraph Problems\thanks{This work was supported by a pile of money from a funding agency.}

### 2017-02-16 paper/paper.tex
This commit might make things worse, but I think we should say something along these lines and make it more coherent tomorrow

Before discussing the algorithm in detail, we start with an illustration of the
main concepts using the graphs $G$ and $H$ in \cref{fig:alg1}.  These graphs
have a maximum common subgraph with four vertices; one example is the mapping
$\{1a, 2f, 3d, 5b\}$ where vertex $1$ is assigned to vertex $a$, $2$ is
assigned to $f$, $3$ to $d$ and $5$ to $b$.

@@

\paragraph{Efficient representation of label classes} The algorithm as
described so far apparently requires $\BigO{g(g+h)}$ space at each level of the
search tree.  We can reduce the memory requirement to $\BigO{g+h}$ per level by
storing a \emph{label class} as a pair $\langle P,T \rangle$ for each label $l$
that is used, where $P$ is the set of vertices in $\V(G)$ labelled $l$, and $T$
is the set of vertices in $\V(H)$ labelled $l$\footnote{We adopt terminology
from subgraph isomorphism, where one graph is a pattern graph $P$ and the other
the target graph $T$.}. Since there are $g + h$ vertices in the two graphs, at
most $g + h$ label classes can exist at once, and there are at most $g + h$
vertices in the union of all of the $P$ and $T$ sets.  We can gain further
efficiency by discarding during search any label class that has an empty $P$ or
$T$ set.

?? Talk about how we can use a pair of flat arrays and indices to the start of
label classes, and throw in citation to partition backtracking here.

\begin{algorithm}[ht]
\DontPrintSemicolon
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$save(matching)$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\nl $bf \gets 0$ \label{SetBestFutureToZero} \;
\nl \lFor {$\langle P,T \rangle \in \AlgVar{future}$}{$\AlgVar{bf} \gets \AlgVar{bf} + \min(|P|,|T|)$}
\nl \lIf {$|M|  + \AlgVar{bf} \leq |\AlgVar{incumbent}|$}{return} \label{PruneSearch}
\medskip
\nl $\langle P,T \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(P)$ \label{SelectVertex} \;
\nl \For {$w \in T$ \label{WLoop}} {
    \nl    $M \gets M \cup \{(v,w)\}$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle P',T'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $P'' \gets P' \cap \N(G, v) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $T'' \gets T' \cap \N(H, w) \setminus \{w\}$ \;
\nl        \If {$P'' \neq \emptyset$ \bf{and} $T'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle P'' , T'' \rangle$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $P'' \gets P' \cap \invN(G, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $T'' \gets T' \cap \invN(H, w) \setminus \{w\}$ \;
\nl        \If {$P'' \neq \emptyset$ \bf{and} $T'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle P'' , T'' \rangle$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{expand}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $P' \gets P \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle P,T \rangle\}$\;
\nl \lIf {$P' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle P',T \rangle \}$}
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(G,H)$ \label{McSplitFun} \;
\nl \Begin{
\nl $\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{expand}(\{\langle V(G),V(H) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl return $|\AlgVar{incubent}|$ \;
}
\caption{Finding a maximum common subgraph.}
\label{McCliqueAlg}
\end{algorithm}

\paragraph{\cref{McCliqueAlg} Explained} The recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.

>>>>

\paragraph{Walkthrough} Before discussing the algorithm in detail, we start
with an illustration of the main concepts using the graphs $G$ and $H$ in
\cref{fig:alg1}.  These graphs have a maximum common subgraph with four
vertices; one example is the mapping $\{1a, 2f, 3d, 5b\}$ where vertex $1$ is
assigned to vertex $a$, $2$ is assigned to $f$, $3$ to $d$ and $5$ to $b$.

@@

\paragraph{Label classes} We require only $\BigO{g+h}$ space per level of the
search tree to store labelling information.  This is done by storing a
\emph{label class} as a pair $\langle P,T \rangle$ for each label $l$ that is
used, where $P$ is the set of vertices in $\V(G)$ labelled $l$, and $T$ is the
set of vertices in $\V(H)$ labelled $l$\footnote{We adopt terminology from
subgraph isomorphism, where one graph is a pattern graph $P$ and the other the
target graph $T$.}. Since there are $g + h$ vertices in the two graphs, at most
$g + h$ label classes can exist at once, and there are at most $g + h$ vertices
in the union of all of the $P$ and $T$ sets. Furthermore, we do not actually
need to store the bits making up a label---we care only that like-labelled
vertices are kept together, and the label itself is not used. Nor do we need to
store any label class which is present only in one graph but not the other (or
which is not present at all).  Together, these facts allow us to store all the
necessary information in two pairs of flat arrays. One pair of arrays is used
to store a permutation of the vertices in $P$ having like-labelled vertices
stored consecutively, together with the indices of the start of each label in
turn (in some arbitrary order). The second pair stores the permutation of
vertices in $T$, together with label start indices \emph{in the same order} as
they are used in the first pair of arrays. ?? This representation has
similarities to data structures used in partition backtracking and bron
kerbosch.

\begin{algorithm}[t]
\DontPrintSemicolon
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$save(matching)$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\nl $bf \gets 0$ \label{SetBestFutureToZero} \;
\nl \lFor {$\langle P,T \rangle \in \AlgVar{future}$}{$\AlgVar{bf} \gets \AlgVar{bf} + \min(|P|,|T|)$}
\nl \lIf {$|M|  + \AlgVar{bf} \leq |\AlgVar{incumbent}|$}{return} \label{PruneSearch}
\medskip
\nl $\langle P,T \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(P)$ \label{SelectVertex} \;
\nl \For {$w \in T$ \label{WLoop}} {
    \nl    $M \gets M \cup \{(v,w)\}$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle P',T'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $P'' \gets P' \cap \N(G, v) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $T'' \gets T' \cap \N(H, w) \setminus \{w\}$ \;
\nl        \If {$P'' \neq \emptyset$ \bf{and} $T'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle P'' , T'' \rangle$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $P'' \gets P' \cap \invN(G, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $T'' \gets T' \cap \invN(H, w) \setminus \{w\}$ \;
\nl        \If {$P'' \neq \emptyset$ \bf{and} $T'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle P'' , T'' \rangle$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{expand}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $P' \gets P \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle P,T \rangle\}$\;
\nl \lIf {$P' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle P',T \rangle \}$}
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(G,H)$ \label{McSplitFun} \;
\nl \Begin{
\nl $\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{expand}(\{\langle V(G),V(H) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl return $|\AlgVar{incubent}|$ \;
}
\caption{Finding a maximum common subgraph.}
\label{McCliqueAlg}
\end{algorithm}

\paragraph{\cref{McCliqueAlg} in detail} The recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.
